---
title: "Reproducible Research Project 1"
author: "Sam Biermans"
date: "`r format(Sys.Date())`"
output:
  html_document:
    keep_md: true
---

github repo: [SamBiermans/RepData_PeerAssessment1](https://github.com/SamBiermans/RepData_PeerAssessment1)

## Loading and preprocessing the data

```{r}
library("data.table")
library("ggplot2")

if (!file.exists("activity.csv")) {
    unzip("activity.zip")
}
```
## Reading csv Data into Data.Table

```{r}
activityDT <- data.table::fread(input = "activity.csv")
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

```{r}
Total_Steps <- activityDT[, c(lapply(.SD, sum, na.rm = FALSE)), .SDcols = c("steps"), by = .(date)]

head(Total_Steps, 10)
```

2. Make a histogram of the total number of steps taken each day.

```{r}
ggplot(Total_Steps, aes(x = steps)) +
    geom_histogram(fill = "blue", binwidth = 1000) +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```

3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
Total_Steps[, .(Mean_Steps = mean(steps, na.rm = TRUE), Median_Steps = median(steps, na.rm = TRUE))]
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. `type = "l"`) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
IntervalDT <- activityDT[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps"), by = .(interval)]

ggplot(IntervalDT, aes(x = interval, y = steps)) +
    geom_line(color = "blue", size = 1) +
    labs(title = "Avg. Daily Steps", x = "Interval", y = "Avg. Steps per day")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
IntervalDT[steps == max(steps), .(max_interval = interval)]
```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
nrow(activityDT[is.na(steps), ])
```

2. Devise a strategy for filling in all of the missing values in the dataset.

```{r}
# Filling in missing values with the median of the dataset.
activityDT[is.na(steps), "steps"] <- activityDT[, c(lapply(.SD, median, na.rm = TRUE)), .SDcols = c("steps")]
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
data.table::fwrite(x = activityDT, file = "tidyData.csv", quote = FALSE)
```

4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
Total_Steps <- activityDT[, c(lapply(.SD, sum)), .SDcols = c("steps"), by = .(date)]

Total_Steps[, .(Mean_Steps = mean(steps), Median_Steps = median(steps))]

ggplot(Total_Steps, aes(x = steps)) +
    geom_histogram(fill = "blue", binwidth = 1000) +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```

## Are there differences in activity patterns between weekdays and weekends?

```{r}
activityDT <- data.table::fread(input = "activity.csv")
activityDT[, date := as.POSIXct(date, format = "%Y-%m-%d")]
activityDT[, `Day of Week` := weekdays(x = date)]
activityDT[grepl(pattern = "Monday|Tuesday|Wednesday|Thursday|Friday", x = `Day of Week`), "weekday or weekend"] <- "weekday"
activityDT[grepl(pattern = "Saturday|Sunday", x = `Day of Week`), "weekday or weekend"] <- "weekend"
activityDT[, `weekday or weekend` := as.factor(`weekday or weekend`)]
head(activityDT, 10)
```

2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
activityDT[is.na(steps), "steps"] <- activityDT[, c(lapply(.SD, median, na.rm = TRUE)), .SDcols = c("steps")]
IntervalDT <- activityDT[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = c("steps"), by = .(interval, `weekday or weekend`)]

ggplot(IntervalDT, aes(x = interval, y = steps, color = `weekday or weekend`)) +
    geom_line() +
    labs(title = "Avg. Daily Steps by Weektype", x = "Interval", y = "No. of Steps") +
    facet_wrap(~`weekday or weekend`, ncol = 1, nrow = 2)
```
